Random Forests for Automatic Matching Bullet-Lands

### 2015-12-22 imaging-paper.Rnw
re-formatting

lines <- matches %>% group_by(lineid) %>% summarise(
  meany = mean(y, na.rm=T),
  miny = min(y, na.rm=T),
  maxy = max(y, na.rm=T) + 1.5625,
  match = isMatch(lineid, type),
  type = type1[1]
)
lines <- subset(lines, lineid != 0)
@
\begin{figure}[hbtp]
\centering
\begin{subfigure}[t]{\linewidth}\centering
\caption{Cross section at $x = 100$.\label{fig:crosscut}}{%
<<crosscuts, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3, out.width='.5\\textwidth'>>=
qplot(y, resid, group=bullet, colour=bullet, data=lof, 
      geom="line") + 
  scale_colour_manual("", values=cols) + theme_bw() +
  theme(legend.position = c(.9, .2),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab("Residuals of loess fit")
@
    }
\end{subfigure}    
\begin{subfigure}[t]{\linewidth}\centering
    \caption{Adjusted (bullet 1-5 horizontally shifted) cross sections at $x = 100$.\label{fig:crosscutX}}{%
<<crosscutsX, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3, out.width='.5\\textwidth'>>=
# qplot(y, resid, group=bullet, colour=bullet, data=lofX, 
#       geom="line", alpha=I(0.6)) + 
#   scale_colour_brewer(palette="Set1") + theme_bw() +
#   theme(legend.position="none")
qplot(y, resid, group=bullet, colour=bullet, data=lofY, 
      geom="line") + 
  scale_colour_manual("", values=cols) + theme_bw() +
  theme(legend.position="none") + ylab("Residuals of loess fit")
@
    }
\end{subfigure}
\caption{\label{fig:cross100}Lines showing surface cross sections at $x = 100$. A horizontal shift of the values of bullet 1-5 to the right shows the similarity of the striation marks.}
\end{figure}

@@

\subsection{Varying cross sections}
The cross section at which a comparison is done matters -- as cross sections are further apart, more pronounced differences between the cross sections show up. This poses both a caveat to matching attempts as well as an opportunity \hh{for quality control}: on the one hand we have to be aware of which cross section level was used in a matching, but on the other hand we can use multiple cross sections of the same bullet for an initial assessment of its quality. Both of these situations are assessed and discussed in more detail below: 
Figure~\ref{fig:crosscuts} shows a sequence of cross sections for bullet 1-5 (barrel 1) that are taken 50$\mu m$ apart, between 150$\mu m$  and 400$\mu m$. These are compared to a cross section at 100$\mu m$. Initially this comparison constitutes an almost perfect match between the two cross sections. However, the match quickly deteriorates with increasing distance to non-matches at a cross section of $x = 300\mu m$ or higher.  Only if the cross section is within 150$\mu m$ do we get a good visual match even when we know that the same bullet surface is being used. 
Given that we have to expect some variation in nominally the same cross section levels due to manual alignments in microscopes, we should take cross section levels into account in the automatic matching routine by evaluating matches at several cross sections. 
\begin{figure}[hbtp]
  \centering
<<crosscuts-vary, echo=FALSE, fig.width = 7, fig.height = 12, out.width = '.5\\linewidth', cache=TRUE, warning = FALSE>>=
paths <- file.path(datadir, dir(datadir))
crosscuts <- seq(100, 400, by = 50)
lof <- processBullets(paths[5], x = crosscuts)
lof$bullet <- paste(lof$bullet, lof$x)

@@

grid.arrange(lop[[1]], lop[[2]], lop[[3]], lop[[4]], lop[[5]], lop[[6]],
             ncol = 1)
@
\caption{\label{fig:crosscuts}Overview of the variations in the surface along different cross sections. Cross section at $x = 100$ is compared to cross sections every 50$\mu m$. With every step away from the original cross section site, the number of differences between the cross sections increases, and the number of maximum CMS decreases from initially 14 to  3 or fewer at $x = 300\mu m$ and higher. }
\end{figure}
%\afterpage{\clearpage}

@@

\begin{figure}[hbtp]
  \centering
<<crosscuts-vary-b31, echo=FALSE, fig.width = 7, fig.height = 10, out.width = '.5\\linewidth', cache=TRUE, warning = FALSE>>=
paths <- file.path(datadir, dir(datadir))
crosscuts <- seq(100,225, by=25)
lof <- processBullets(paths[37], x = crosscuts)
lof$bullet <- paste(lof$bullet, lof$x)

@@

grid.arrange(lop[[1]], lop[[2]], lop[[3]], 
             lop[[4]], lop[[5]], 
             ncol=1)
@
\caption{\label{fig:crosscuts2}Varying cross sections for barrel 3, bullet 1. Initially, the match between cross sections 25$\mu m$ apart is affected strongly by break off at the bottom of the bullet. At a cross section level of $175\mu m$ the bullet pattern stabilizes. For this land, matches should not be attempted at lower cross sections.}
\end{figure}
An opportunity that arises from comparing multiple cross sections of a bullet's surface measurements is the following: by comparing cross sections that are not too far apart -- maybe 25 or 50 $\mu m$ for example -- we get an indication whether the cross section is in a rapidly changing section of the surface, indicative of a break-off \hh{or some other damage}, or in a stable section, where we can expect to find matches to other surfaces reasonably well. In the approach here, we keep increasing the cross section level until we find a section with a stable pattern. This process is shown in Figure~\ref{fig:crosscuts2} at the example of bullet 1-1 from barrel 3.
`Stability' is here defined as two aligned cross sections chosen 25$\mu m$ apart having a cross-correlation of at least 0.9.

@@

\begin{figure}[hbtp]
  \centering
<<barrel2, echo=FALSE, fig.width = 7, fig.height=12, out.width='.5\\linewidth', warning=FALSE>>=

@@

grid.arrange(list_of_matches[[1]], list_of_matches[[2]], list_of_matches[[3]], 
             list_of_matches[[4]], list_of_matches[[5]], list_of_matches[[6]],
             ncol = 1)
@
\caption{\label{fig:barrel2}All lands from the first bullet shot from barrel 2 are compared to the first land of the second bullet shot from barrel 2. The largest number of CMS is eleven, observed with land six of the first bullet. }
\end{figure}

@@

qplot(cmsdist, geom="bar") + theme_bw() + xlab("Number of CMS")
@
\end{subfigure}
\begin{subfigure}[t]{\textwidth}\centering
\caption{Overlaid lines of C-3 and the land with the top matching CMS.}
<<top, echo=FALSE, fig.width=7, fig.height=3.25, out.width='.5\\textwidth', warning = FALSE>>=
res <- reslist[[which.max(cmsdist)]]  
#res <- reslist[[which.max(cmsdist[-which.max(cmsdist)])]]  # number 2
res$bullets$bullet <- scrubPath(res$bullets$bullet)

@@

@
\end{subfigure}
\begin{subfigure}[t]{\textwidth}\centering
\caption{Overlaid lines of C-3 and the land with the second top matching CMS.}
<<top2, echo=FALSE, fig.width=7, fig.height=3.25, out.width='.5\\textwidth', warning = FALSE>>=
res <- reslist[[which(cmsdist==14)]]  # number 2
res$bullets$bullet <- scrubPath(res$bullets$bullet)

@@

Another benefit is that we can not only apply a single decision tree, but several hundreds of them to combine in a random forest~\citep{breiman:2001, randomForest}. In a random forest multiple trees are fitted. For each of these trees, only two thirds of the observations are used for fitting, while predictions for the remaining third  are used to evaluate the tree's predictive power and accuracy, or its reverse, the error rate. Because errors are determined from the one third of held-back observations, this error rate is called the out-of bag (OOB) error. 
Figure~\ref{fig:oob} shows the cumulative out-of-bag error (OOB) rate for 300 trees. 
%
\begin{figure}[hbtp]
  \centering
<<oob, echo=FALSE, fig.width=7, fig.height=3, out.width='.5\\linewidth'>>=
qplot(id, OOB, geom="line", data=errors) + theme_bw() + ylim(c(0,NA)) + xlab("Number of trees") + ylab("Out of Bag Error (OOB)")
@
\caption{\label{fig:oob}Cumulative out-of-bag error rate of a random forest fit to predict land-to-land matches from image features.}
\end{figure}
%
After about 25 trees, the error rate of land-to-land comparisons stabilizes at 0.0040. This is a weighted average between a false positive error rate of 0.000098 and an error rate of false negatives of 0.22. \hh{This out-of-bag error rate is over-estimating the actual error in the Hamby study: here, the final random forest based on 300 trees is able to correctly predict all known matches and non-matches (see Figures~\ref{fig:treeresult} and ~\ref{fig:tree-forest}).}
Note that this error rate is based on land-to-land comparisons and is much lower for bullet-to-bullet matches. In the case of the Hamby study, the overall error rate is zero, if we require for a bullet match that at least two of a bullet's lands are matched (see Figure~\ref{fig:treeresult}). 
\begin{figure}[hbtp]
  \centering
<<overall-tree, echo=FALSE, fig.width=7, fig.height=4, out.width='.5\\linewidth'>>=
bstats$bullet <- gsub("-[0-9]$", "", bstats$b2)
bullets <- bstats %>% filter(match, !flagged) %>% group_by(bullet) %>% summarize(
  n = n(),
  pred = sum(pred > 0.5),
  forest = sum(forest > 0.5)
)
library(reshape2)
bm <- melt(data=bullets, id.var=c("bullet", "n"))

@@

For a fair assessment of the adequacy of an algorithm we need transparency. Our matching  algorithm is open: the code is readily available in form of the R package x3prplus~\citep{x3prplus}, and the code to produce this paper is available at \url{http://www.github.com/erichare/imaging-paper}. Obviously, this is only a first step, but in times of the Netflix prize~\citep{koren:2009, toescher:2009, piotte:2009} and the many Data Science challenges hosted by Kaggle\footnote{kaggle home:\url{https://www.kaggle.com/solutions/competitions}}, we should not let the quality of the matching algorithm be the problem that stands in the way.

>>>>

lines <- matches %>% group_by(lineid) %>% summarise(
  meany = mean(y, na.rm=T),
  miny = min(y, na.rm=T),
  maxy = max(y, na.rm=T) + 1.5625,
  match = isMatch(lineid, type),
  type = type1[1]
)
lines <- subset(lines, lineid != 0)
@
\begin{figure}[hbtp]
\centering
\begin{subfigure}[b]{.49\linewidth}\centering
\caption{Cross section at $x = 100$.\label{fig:crosscut}}{%
<<crosscuts, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3, out.width='\\textwidth'>>=
qplot(y, resid, group=bullet, colour=bullet, data=lof, 
      geom="line") + 
  scale_colour_manual("", values=cols) + theme_bw() +
  theme(legend.position = c(.9, .2),
        legend.background = element_rect(fill=alpha('white', 0.4))) + 
  ylab("Residuals of loess fit")
@
    }
\end{subfigure}    
\begin{subfigure}[b]{.49\linewidth}\centering
    \caption{Adjusted (bullet 1-5 horizontally shifted) cross sections at $x = 100$.\label{fig:crosscutX}}{%
<<crosscutsX, echo=FALSE, dependson='twolands100', fig.width=7, fig.height=3, out.width='\\textwidth'>>=
# qplot(y, resid, group=bullet, colour=bullet, data=lofX, 
#       geom="line", alpha=I(0.6)) + 
#   scale_colour_brewer(palette="Set1") + theme_bw() +
#   theme(legend.position="none")
qplot(y+min(lof$y), resid, group=bullet, colour=bullet, data=lofY, 
      geom="line") + xlab("y") +
  scale_colour_manual("", values=cols) + theme_bw() +
  theme(legend.position="none") + ylab("Residuals of loess fit")
@
    }
\end{subfigure}
\caption{\label{fig:cross100}Lines showing surface cross sections at $x = 100$. A horizontal shift of the values of bullet 1-5 to the right shows the similarity of the striation marks.}
\end{figure}

@@

\subsection{Varying cross sections}
The cross section at which a comparison is done matters -- as cross sections are further apart, more pronounced differences between the cross sections show up. This poses both a caveat to matching attempts as well as an opportunity \hh{for quality control}: on the one hand we have to be aware of which cross section level was used in a matching, but on the other hand we can use multiple cross sections of the same bullet for an initial assessment of its quality. Both of these situations are assessed and discussed in more detail below: 
Figure~\ref{fig:crosscuts} shows a sequence of cross sections for bullet 1-5 (barrel 1) that are taken 50$\mu m$ apart, between 150$\mu m$  and 400$\mu m$. These are compared to a cross section at 100$\mu m$. Initially this comparison constitutes an almost perfect match between the two cross sections. However, the match quickly deteriorates with increasing distance to non-matches at a cross section of $x = 300\mu m$ or higher.  Only if the cross section is within 150$\mu m$ do we get a good visual match even when we know that the same bullet surface is being used. 
Given that we have to expect some variation in nominally the same cross section levels due to manual alignments in microscopes, we should take cross section levels into account in the automatic matching routine by evaluating matches at several cross sections. 
\begin{figure}[hbtp]
  \centering
<<crosscuts-vary, echo=FALSE, fig.width = 12, fig.height = 7, out.width = '\\linewidth', cache=TRUE, warning = FALSE>>=
paths <- file.path(datadir, dir(datadir))
crosscuts <- seq(100, 400, by = 50)
lof <- processBullets(paths[5], x = crosscuts)
lof$bullet <- paste(lof$bullet, lof$x)

@@

grid.arrange(lop[[1]], lop[[2]], lop[[3]], lop[[4]], lop[[5]], lop[[6]],
             ncol = 2)
@
\caption{\label{fig:crosscuts}Overview of the variations in the surface along different cross sections. Cross section at $x = 100$ is compared to cross sections every 50$\mu m$. With every step away from the original cross section site, the number of differences between the cross sections increases, and the number of maximum CMS decreases from initially 14 to  3 or fewer at $x = 300\mu m$ and higher. }
\end{figure}
%\afterpage{\clearpage}

@@

\begin{figure}[hbtp]
  \centering
<<crosscuts-vary-b31, echo=FALSE, fig.width = 12, fig.height = 7, out.width = '\\linewidth', cache=TRUE, warning = FALSE>>=
paths <- file.path(datadir, dir(datadir))
crosscuts <- seq(100,250, by=25)
lof <- processBullets(paths[37], x = crosscuts)
lof$bullet <- paste(lof$bullet, lof$x)

@@

grid.arrange(lop[[1]], lop[[2]], lop[[3]], 
             lop[[4]], lop[[5]], lop[[6]],
             ncol=2)
@
\caption{\label{fig:crosscuts2}Varying cross sections for barrel 3, bullet 1. Initially, the match between cross sections 25$\mu m$ apart is affected strongly by break off at the bottom of the bullet. At a cross section level of $175\mu m$ the bullet pattern stabilizes. For this land, matches should not be attempted at lower cross sections.}
\end{figure}
An opportunity that arises from comparing multiple cross sections of a bullet's surface measurements is the following: by comparing cross sections that are not too far apart -- maybe 25 or 50 $\mu m$ for example -- we get an indication whether the cross section is in a rapidly changing section of the surface, indicative of a break-off \hh{or some other damage}, or in a stable section, where we can expect to find matches to other surfaces reasonably well. In the approach here, we keep increasing the cross section level until we find a section with a stable pattern. This process is shown in Figure~\ref{fig:crosscuts2} at the example of bullet 1-1 from barrel 3.
`Stability' is here defined as two aligned cross sections chosen 25$\mu m$ apart having a cross-correlation of at least 0.9.

@@

\begin{figure}[hbtp]
  \centering
<<barrel2, echo=FALSE, fig.width = 10, fig.height=6, out.width='\\linewidth', warning=FALSE>>=

@@

grid.arrange(list_of_matches[[1]], list_of_matches[[2]], list_of_matches[[3]], 
             list_of_matches[[4]], list_of_matches[[5]], list_of_matches[[6]],
             ncol = 2)
@
\caption{\label{fig:barrel2}All lands from the first bullet shot from barrel 2 are compared to the first land of the second bullet shot from barrel 2. The largest number of CMS is eleven, observed with land six of the first bullet. }
\end{figure}

@@

qplot(cmsdist, geom="bar") + theme_bw() + xlab("Number of CMS")
@
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Overlaid lines of C-3 and the land with the top matching CMS.}
<<top, echo=FALSE, fig.width=7, fig.height=3.25, out.width='\\textwidth', warning = FALSE>>=
res <- reslist[[which.max(cmsdist)]]  
#res <- reslist[[which.max(cmsdist[-which.max(cmsdist)])]]  # number 2
res$bullets$bullet <- scrubPath(res$bullets$bullet)

@@

@
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Overlaid lines of C-3 and the land with the second top matching CMS.}
<<top2, echo=FALSE, fig.width=7, fig.height=3.25, out.width='\\textwidth', warning = FALSE>>=
res <- reslist[[which(cmsdist==14)]]  # number 2
res$bullets$bullet <- scrubPath(res$bullets$bullet)

@@

Another benefit is that we can not only apply a single decision tree, but several hundreds of them to combine in a random forest~\citep{breiman:2001, randomForest}. In a random forest multiple trees are fitted. For each of these trees, only two thirds of the observations are used for fitting, while predictions for the remaining third  are used to evaluate the tree's predictive power and accuracy, or its reverse, the error rate. Because errors are determined from the one third of held-back observations, this error rate is called the out-of bag (OOB) error. 
Figure~\ref{fig:oob} shows the cumulative out-of-bag error (OOB) rate for 300 trees. 
%
\begin{figure}[hbtp]
  \centering
<<oob, echo=FALSE, fig.width=7, fig.height=3, out.width='.5\\linewidth'>>=
qplot(id, OOB, geom="line", data=errors) + theme_bw() + ylim(c(0,NA)) + xlab("Number of trees") + ylab("Out of Bag Error (OOB)")
@
\caption{\label{fig:oob}Cumulative out-of-bag error rate of a random forest fit to predict land-to-land matches from image features.}
\end{figure}
%
After about 25 trees, the error rate of land-to-land comparisons stabilizes at 0.0040. This is a weighted average between a false positive error rate of $9.8 \times 10^{-5}$ and an error rate of false negatives of 0.22. \hh{This out-of-bag error rate is over-estimating the actual error in the Hamby study: here, the final random forest based on 300 trees is able to correctly predict all known matches and non-matches (see Figures~\ref{fig:treeresult} and ~\ref{fig:tree-forest}).}
Note that this error rate is based on land-to-land comparisons and is much lower for bullet-to-bullet matches. In the case of the Hamby study, the overall error rate is zero, if we require for a bullet match that at least two of a bullet's lands are matched (see Figure~\ref{fig:treeresult}). 
\begin{figure}[hbtp]
  \centering
<<overall-tree, echo=FALSE, fig.width=7, fig.height=4, out.width='.5\\linewidth'>>=
bstats$bullet <- gsub("-[0-9]$", "", bstats$b2)
bullets <- bstats %>% filter(match, !flagged) %>% group_by(bullet) %>% summarize(
  n = n(),
  pred = sum(pred > 0.5),
  forest = sum(forest > 0.5)
)
library(reshape2)
bm <- melt(data=bullets, id.var=c("bullet", "n"))

@@

For a fair assessment of the adequacy of an algorithm we need transparency. Our matching  algorithm is open: the code is readily available in form of the R package x3prplus~\citep{x3prplus}, and the code to produce this paper is available at \url{http://www.github.com/erichare/imaging-paper}. Obviously, this is only a first step, but in times of the Netflix prize~\citep{koren:2009, toescher:2009, piotte:2009} and the many Data Science challenges hosted by Kaggle\footnote{kaggle home: \url{https://www.kaggle.com/solutions/competitions}}, we should not let the quality of the matching algorithm be the problem that stands in the way.
